Consistency scores in text data

### 2020-08-03 outputs/documents/paper.Rmd
Update first four pages

---
output: 
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: false
    toc: false
    fig_caption: true
    latex_engine: pdflatex
    template: templates/svm-latex-ms.tex
bibliography: "references.bib"
header-includes:
  -  \usepackage{hyperref}
biblio-style: apalike
title: "Consistency scores in text data"
thanks: "Thank you to X, Y and Z for helpful comments. A Shiny application for interactive and small-scale examples is available: https://kelichiu.shinyapps.io/Arianna/. An R package for larger scale applications is available: **HERE**. Our code and datasets are available: https://github.com/RohanAlexander/consistency_scores_in_text_datasets. Comments on the `r format(Sys.time(), '%d %B %Y')` version of this paper are welcome at: rohan.alexander@utoronto.ca."
author:
- name: Rohan Alexander
  affiliation: University of Toronto
- name: Ke-Li Chiu
  affiliation: University of Toronto  
abstract: "In this paper we introduce a process to clean the text extracted from PDFs using various methods from natural language processing. Our approach compares originally extracted text with the text generated from, or expected by, these methods using earlier text as stimulus. To guide this process, we introduce the notion of a consistency score, which refers to the proportion of text that is unchanged by the model. This is used to monitor changes during the cleaning process, and to compare the messiness of different texts. The methods that we consider are: n-grams, [Google’s Word2vec, Standford's GloVe], GPT-2, and GPT-3. We illustrate our process on text from the Canadian Hansard and introduce both a Shiny application and an R package to make our process easier for others to adopt."
keywords: "text-as-data; natural language processing; GPT-3; quantitative analysis."
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 12pt
# spacing: double
endnote: no
graphics: yes
---

@@

In this paper we introduce the concept of consistency in a text corpus. Consistency refers to the proportion of words that are able to be forecast by a statistical model, based on preceeding words and surrounding context. Further, we define internal consistency as when the model is trained on the corpus itself, and external consistency as when the model is trained on a more general corpus. Together, these concepts provide a guide to the cleanliness and consistency of a text dataset. This can be important when deciding whether a dataset is fit for purpose; when carrying out data cleaning and preparation tasks; and as a comparison between datasets.

To provide an example, consider the sentence, 'the cat in the\dots'. A child who has read this book could tell you that the next word should be 'hat'. Hence if the sentence was actually 'the cat in the bat', then that child would know something was likely wrong. The consistency score would likely be lower than if the sentence were 'the cat in the hat'. After we correct this error, the consistency score would likely increase. By examining how the consistency scores evolve in response to changes made to the text during the data preparation and cleaning stages we can better understand the effect of the changes. Including consistency scores when text corpuses are shared allows researchers to be more transparent about their corpus. And finally, the use of consistency scores allows for an increased level of automation in the cleaning process.

We consider various natural language processing methods. These include n-grams, [Google’s Word2vec, Standford's GloVe], as well as GPT2 and GPT-3. The n-gram approach involves identifying two, three, or more, words that are commonly found together. Here think of a two-gram involving the word 'good' 'good morning'. At scale these can identify missing or unusual words, and work quickly, but they lack nuance. For instance an equally reasonable two-gram involving the word 'good' is 'good work'. For that reason we consider pre-trained word embedding models. These include [Google’s Word2vec, Standford's GloVe], which works [representing words as dense vectors that capture their semantic relationships with each other]**SOMEHOW**. We also consider pre-trained generative models such as GPT-2 and GPT-3 from OpenAI. GPT-2 and GPT-3 are pre-trained generative unsupervised language models. They differ in the number of parameters. Generative Pretrained Transformer (GPT)-2 has *X* parameters and GPT-3 has more than 175 billions parameters. While GPT-2 is more limited, it can be run on smaller machines, whereas GPT-3 cannot.

@@

Without exception, all human languages have some words that co-occur more frequently with others. Under this assumption, we can regard the production of English text as a set of conditional probabilities, written as Pr($w_{k}$ | $w^{k-1}_{1}$), where k is the number of words in a sentence, $w_{k}$ is the predicted word, and $w^{k-1}_{1}$ is the history of the word occurring in a sequence (@brown1992class). In other word, the generation of prediction $w_{k}$ is based on the history $w^{k-1}_{1}$; this is the overarching principle of all language models. Essentially, a language model is the probability distribution over sequences of words **(Ke-Li add ref)**. The goal of statistical language modeling is to estimate the probability distribution of different linguistic units — words, sentences, and even documents **(Ke-Li add ref)**. However, this is fundamentally difficult because language is categorical. If we see each word in a set of vocabulary as a category, the dimensions of language is extremely high (@rosenfeld2000two). In the course of statistical language model development, [how n-gram, word embedding, transformers tackle this].

@@

PDFs contain a lot of information, but typically that information is not able to be analysed directly by statistical models. It needs to first be extracted from the PDF, often by optical character recognition (OCR), and then cleaned, prepared and made into some type of tabular dataset. One issue in this process is that OCR is not perfect. Hence there is a need to correct the output before it can be analysed. Traditionally, word-correction techniques evaluate errors one by one without considering the context of the surrounding words. This was no longer the case in modern correction techniques as statistical language models (SLMs) and feature-based methods have been used for context-sensitive correction (citation).

This conditional probability of the language model is the foundation of an n-gram language model. An n-gram model is a probabilistic language model that predicts the next word in a sequence of words. The $n$ in n-gram represents the number of words in a sequence. Taking an incomplete sentence from the Jane Eyre excerpt as an example:

> "We had been wandering,"

"We" is a uni-gram, "We had" is a bi-gram, "We had been" is a tri-gram, and "We had been wandering" is a 4-gram. On the other hand, the tri-grams of the excerpt are "We had been," and "had been wandering." The overlapping chaining of words reveals the statistical behaviour in human language. Moreover, using n-gram models then enable us to assign a probability to the occurrence of a sequence of words or the likelihood of a word occurring next. Consider the two sentences: "We had been wandering," and "We had been wangling." The former is likely to be more frequently encountered in a training corpus. Thus, the n-gram would assign a higher probability to "wandering" than to "wangling."

To predict which word appears next, we have to take the sequence of preceeding words into account, which requires knowing the probability of the sequence of words. The probability of a sequence appearing in a corpus follows the chain rule:

> P(We,had,been,wandering) $=$ P(We)$\cdot$P(had|We)$\cdot$P(been|We,had)$\cdot$P(wandering|We,had,been)

However, the likelihood that more and more words will occur next to each other in an identical sequence becomes smaller and smaller, making the prediction difficult. Alternatively, we can approximate the probability of a word depending on only the previous word; this is known as Markov assumption. Instead, the Markov assumption permits us to approximate the probability using only the last $n$ words (@brown1992class). More specifically:

> P(We,had,been,wandering)$\approx$ P(We)$\cdot$P(had|We)$\cdot$P(been|had)$\cdot$P(wandering|been)

Because the bi-gram model looks only the immediate preceding word, an n-gram model under Markov assumption can then be reduced into a bi-gram model with n being any number. The general equation is:

> P($w_{n} | w^{n-1}_{1}) \approx$ P($w_{n} | w^{n-1}_{n - N+1})$)

The application of n-gram is versatile. N-gram models are widely applied in text prediction, spelling-correction, machine translation **(Keli add citation)**. In our application, we use a tri-gram based model to detect both real-word and non-word errors and correct them with the candidate word that has the highest probability. [add more about how we use n-gram once the functions in the app are in place]

The n-gram models work do not take the linguistic structure of language into account. As @rosenfeld2000two pointed out, "... it may as well be a sequence of arbitrary symbols, with no deep structure, intention or thought behind". The prediction of the next word is only from a few words that are immediately preceding; the context is not taking into account. For this shortcoming, the predictive power of n-gram is largely limited.

@@

Developed by Google, Word2Vec is a word embedding model that uses shallow neural network. While it is not a deep learning model, it turns words into a vector representations that deep neural networks can understand. It takes a corpus as its input and output a set of feature vectors that represent the words in the given corpus. @mikolov2013efficient proposed two architectures for learning distributed representations of words within the framework of Word2Vec — Continuous Bag-of-Words Model (CBOW) and Continuous Skip-gram Model. Both architectures contains three layers: input layer, projection layer and output layer. The input layer encodes words within a context window to one-hot vectors. Using a projection matrix, the input layer is  projected to a projection layer and the features are. Finally, the output layer is a softmax regression classifier that obtains the posterior distribution of words (@rong2014word2vec). The input and output vectors differ in the two architecutres. CBOW uses several surrounding words, or the context, to predict a target word. On the other hand, continuous skip-gram takes a single word as input to predict its neighbouring words. For example, consider the sentence "I think therefore I am", a famous quote by Descartes. The focus word is we want to learn the embedding is set to "therefore", and the context window is set to two which will give us the two immediate preceding words and two immediate proceeding words of the focus word. The input for the CBOW is "I, think, I, am" and the output is "therefore". In contrast, the skip-gram model will take "therefore" as the input and out "I, think, I, am" as the prediction.

@@

The ratio between the co-occurrence probability of "ice, solid" pair and "steam, solid" pair is high because the co-occurrence of "ice" and "solid" is higher than the co-occurrence of "steam" and "solid". Reversely, the ratio between the co-occurrence of "ice, gas" pair and "steam, gas" pair is low. Other other hand, the ratio of co-occurrence probability of "ice, water" pair and "steam, water" pair are close to 1 because water is relevant to both "ice" and "steam". Probability alone does not distinguish well relevant words from irrelevant words. For this reason, the researcher of GloVe argued that "the appropriate starting point for word vector learning should be with ratios of co-occurrence probabilities rather than the probabilities themselves (@pennington2014glove)."

@@

OpenAI has made GPT-2 open-sourced through Github with Python and Tensorflow. There is also a package in R that wraps the Python code into a R package, created by **(Keli add citation)**. However, the R package only permits exploratory experiments with the pre-trained tasks; fine-tuning for new tasks is not available thorough R.

>>>>

---
output: 
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: false
    toc: false
    fig_caption: true
    latex_engine: pdflatex
    template: templates/svm-latex-ms.tex
bibliography: "references.bib"
header-includes:
  -  \usepackage{hyperref}
biblio-style: apalike
title: "Consistency scores in text data"
thanks: "Thank you to X, Y and Z for helpful comments. A Shiny application for interactive and small-scale examples is available: https://kelichiu.shinyapps.io/Arianna/. An R package for larger scale applications is available: **HERE**. Our code and datasets are available: https://github.com/RohanAlexander/consistency_scores_in_text_datasets. Comments on the `r format(Sys.time(), '%d %B %Y')` version of this paper are welcome at: rohan.alexander@utoronto.ca."
author:
- name: Ke-Li Chiu
  affiliation: University of Toronto
- name: Rohan Alexander
  affiliation: University of Toronto
abstract: "In this paper we introduce a process to clean the text extracted from PDFs using various methods from natural language processing. Our approach compares originally extracted text with the text generated from, or expected by, these methods using earlier text as stimulus. To guide this process, we introduce the notion of a consistency score, which refers to the proportion of text that is unchanged by the model. This is used to monitor changes during the cleaning process, and to compare the messiness of different texts. The methods that we consider are: n-grams, Google's Word2vec, Stanford's GloVe, and OpenAI's GPT-2 and GPT-3. We illustrate our process on text from the Canadian Hansard and introduce both a Shiny application and an R package to make our process easier for others to adopt."
keywords: "text-as-data; natural language processing; quantitative analysis."
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 12pt
# spacing: double
endnote: no
graphics: yes
---

@@

In this paper we introduce the concept of consistency in a text corpus. Consistency refers to the proportion of words that are able to be forecast by a statistical model, based on preceding words and surrounding context. Further, we define internal consistency as when the model is trained on the corpus itself, and external consistency as when the model is trained on a more general corpus. Together, these concepts provide a guide to the cleanliness and consistency of a text dataset. This can be important when deciding whether a dataset is fit for purpose; when carrying out data cleaning and preparation tasks; and as a comparison between datasets.

To provide an example, consider the sentence, 'the cat in the\dots'. A child who has read this book could tell you that the next word should be 'hat'. Hence if the sentence was actually 'the cat in the bat', then that child would know something was likely wrong. The consistency score would likely be lower than if the sentence were 'the cat in the hat'. After we correct this error, the consistency score would likely increase. By examining how the consistency scores evolve in response to changes made to the text during the data preparation and cleaning stages we can better understand the effect of the changes. Including consistency scores and their evolution when text corpuses are shared allows researchers to be more transparent about their corpus. And finally, the use of consistency scores allows for an increased level of automation in the cleaning process.

We consider various natural language processing methods. These include n-grams, Google's Word2vec, Stanford's GloVe, and OpenAI's GPT-2 and GPT-3. The n-gram approach involves identifying two, three, or more, words that are commonly found together. Here think of a two-gram involving the word 'good' 'good morning'. At scale these can identify missing or unusual words, and work quickly, but they lack nuance. For instance an equally reasonable two-gram involving the word 'good' is 'good work'. For that reason we consider pre-trained word embedding models. These include Google's Word2vec and Stanford's GloVe, which place each word in a multi-dimensional space such that distance between words can illustrate their relationship. We also consider pre-trained generative models such as OpenAI's GPT-2 and GPT-3. GPT-2 and GPT-3 are pre-trained generative unsupervised language models. They differ in the number of parameters. Generative Pretrained Transformer (GPT)-2 has 1.5 billion parameters and GPT-3 has more than 175 billion parameters. While GPT-2 is more limited, it can be run on smaller machines, whereas GPT-3 cannot and is accessed via an API.

@@

Given the nature of human languages, some combinations of words tend to occur more frequently than others. Think of 'good', which is more often followed by 'morning', than 'duck'. As such, we could consider English text production as a conditional probability, $\mbox{Pr}(w_{k} | w^{k-1}_{1})$, where $k$ is the number of words in a sentence, $w_{k}$ is the predicted word, and $w^{k-1}_{1}$ is the history of the word occurring in a sequence [@brown1992class]. In this way, the generation of some prediction, $w_{k}$, is based on the history, $w^{k-1}_{1}$. This is the underlying principle of all language models. Essentially, a language model is a probability distribution over sequences of words **(Ke-Li add ref)**. The goal of statistical language modeling is to estimate probability distributions over different linguistic units — words, sentences, and even documents **(Ke-Li add ref)**. However, this is difficult as language is categorical. If we consider each word in a vocabulary as a category, then the dimensionality of language becomes large [@rosenfeld2000two]. The reason that there is scuh a variety of statistical language models is that there are various ways of dealing with this fundamental problem.

@@

The Portable Document Format (PDF) is typically used to render documents in a relatively  fixed way. PDFs contain a lot of information, but often that information is not able to be analysed directly by statistical language models. The information first needs to be extracted from the PDF, often by optical character recognition (OCR), and then cleaned, prepared, and made into some type of dataset. One issue in this process is that OCR is not perfect. Hence there is a need to correct the output before it is analysed. One way to do this is to correct errors one-by-one and independent of context. We all know that 'tbe' is wrong, and should be 'the'. But statistical language models have the advantage of correction that can occur within context. This allows significant improvements, for instance sometimes 'bat' should be 'hat' and sometimes it really should be 'bat' **(Ke-Li add ref)**.

The foundation of an n-gram language model is the conditional probability set-up introduced above. An n-gram model is a probabilistic language model that predicts the next word in a sequence of words **(Ke-Li add ref)**. The $n$ in n-gram referes to the number of words in that sequence. Consider the following excerpt from *Jane Eyre*:

> 'We had been wandering'.

'We' is a uni-gram, 'We had' is a bi-gram, 'We had been' is a tri-gram, and 'We had been wandering' is a 4-gram. Notice that the two tri-grams in this excerpt, 'We had been', and 'had been wandering', overlap. The use of n-gram models enables us to assign probabilities to both the next sequence of words and just the next word. For instance, consider the two sentences: 'We had been wandering', and 'We had been wangling'. The former is likely to be more frequently encountered in a training corpus. Thus, an n-gram would assign a higher probability to the next word being 'wandering' than 'wangling', given the sequence 'We had been'.

To predict the next word, we have to take the sequence of preceding words into account, which requires knowing the probability of the sequence of words. The probability of a sequence appearing in a corpus follows the chain rule:

$$\mbox{Pr(We,had,been,wandering)} = \mbox{Pr(We)} \times \mbox{P(had|We)} \times \mbox{P(been|We,had)} \\ 
\times \mbox{P(wandering|We,had,been)}.$$

However, the likelihood that more and more words will occur next to each other in an identical sequence becomes smaller and smaller, making prediction difficult. Alternatively, we can approximate the probability of a word depending on only the previous word. This is known as the 'Markov assumption' and it allows us to approximate the probability using only the last $n$ words [@brown1992class]:

$$\mbox{Pr(We,had,been,wandering)} \approx \mbox{Pr(We)} \times \mbox{P(had|We)} \times \mbox{P(been|had)} \\ 
\times \mbox{P(wandering|been)}.$$

As a bi-gram model only considers the immediately preceding word, under the Markov assumption, an $n$-gram model can be reduced to a bi-gram model with $n$ being any number:

$$\mbox{Pr}(w_{n} | w^{n-1}_{1}) \approx \mbox{P}(w_{n} | w^{n-1}_{n - N+1}))$$
where $N$ is **(Ke-Li add)**.

Language models underpinned by n-grams are widely applied in text prediction, spelling-correction, and machine translation **(Ke-Li add citation)**. In our application, we use a tri-gram based model to detect both real-word and non-word errors and correct them with the candidate word that has the highest probability. **(TODO: Add more about how we use n-gram once the functions in the app are in place.)**

However, n-gram models do not take the linguistic structure of language into account. For instance, @rosenfeld2000two [p. X]  **(Ke-Li add)** discusses language in this context, saying that '...it may as well be a sequence of arbitrary symbols, with no deep structure, intention or thought behind'. The prediction of the next word is based on only a few preceding words and broader context is not taken into account. Hence next-word prediction using n-gram based language models can be limited.

@@

Developed by Google, Word2Vec is a word embedding model that uses shallow neural network. While it is not a deep learning model, it turns words into a vector representations that deep neural networks can understand. It takes a corpus as its input and output a set of feature vectors that represent the words in the given corpus. @mikolov2013efficient proposed two architectures for learning distributed representations of words within the framework of Word2Vec — Continuous Bag-of-Words Model (CBOW) and Continuous Skip-gram Model. Both architectures contains three layers: input layer, projection layer and output layer. The input layer encodes words within a context window to one-hot vectors. Using a projection matrix, the input layer is  projected to a projection layer and the features are. Finally, the output layer is a softmax regression classifier that obtains the posterior distribution of words (@rong2014word2vec). The input and output vectors differ in the two architectures. CBOW uses several surrounding words, or the context, to predict a target word. On the other hand, continuous skip-gram takes a single word as input to predict its neighbouring words. For example, consider the sentence "I think therefore I am", a famous quote by Descartes. The focus word is we want to learn the embedding is set to "therefore", and the context window is set to two which will give us the two immediately preceding words and two immediate proceeding words of the focus word. The input for the CBOW is "I, think, I, am" and the output is "therefore". In contrast, the skip-gram model will take "therefore" as the input and out "I, think, I, am" as the prediction.

@@

The ratio between the co-occurrence probability of "ice, solid" pair and "steam, solid" pair is high because the co-occurrence of "ice" and "solid" is higher than the co-occurrence of "steam" and "solid". Reversely, the ratio between the co-occurrence of "ice, gas" pair and "steam, gas" pair is low. On the other hand, the ratio of co-occurrence probability of "ice, water" pair and "steam, water" pair are close to 1 because water is relevant to both "ice" and "steam". Probability alone does not distinguish well relevant words from irrelevant words. For this reason, the researcher of GloVe argued that "the appropriate starting point for word vector learning should be with ratios of co-occurrence probabilities rather than the probabilities themselves (@pennington2014glove)."

@@

OpenAI has made GPT-2 open-sourced through GitHub with Python and TensorFlow. There is also a package in R that wraps the Python code into a R package, created by **(Keli add citation)**. However, the R package only permits exploratory experiments with the pre-trained tasks; fine-tuning for new tasks is not available thorough R.
